\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-06 jasa/paper-jasa.tex
Added Juan affiliation
Update first example

\if1\blind
{
  \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
  \author{Kyle Cranmer, Juan Pavez and Gilles Louppe\\
          New York University}
  \maketitle
} \fi

@@

As outlined above, any consistent probabilistic classification algorithm can be
used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
expected to smoothly vary around $\mathbf{x}$, we would however recommend
learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
around $\mathbf{x}$, such as neural networks. For small training sets, tree-based
methods are not expected to work so well for this use case, since they usually
model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
In such cases where $\hat s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased
approximation of the density ratio.
Provided enough training data, accurate results could however still be achieved,
given the universal approximator capacity of tree-based models.

@@

As a simple and illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
Figure~\ref{fig:1} and defined as
\begin{align}
p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - g) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - g) p_{c_1}(\mathbf{x}) + g p_{c_2}(\mathbf{x}), \\
p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $g$ is a parameter of $p(\mathbf{x}|\theta_0)$
set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
figure and will be used later for inference.

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:2a}, \ref{fig:2b} and \ref{fig:2c} show the approximated
ratio when using respectively a linear model, a 2-layer neural network and a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when decomposing the mixtures, as
described in Section~\ref{sec:mixture}.
Hyper-parameters of the classifiers are all tuned on a finite grid of values
in order to minimize the mean squared error, as estimated by 3-fold cross-validation.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1.pdf}
        \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
        \label{fig:1}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-a.pdf}
        \caption{Linear model}
        \label{fig:2a}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-b.pdf}
        \caption{Neural network}
        \label{fig:2b}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2-c.pdf}
        \caption{Random forest}
        \label{fig:2c}
    \end{subfigure}
    \caption{Approximation of the log-likelihood ratio
             $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
             with calibrated classifiers.}
\end{figure}

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat
s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
of the ratio, even when using universal approximators such as a neural network
or a random forest. This observation is at the origin of this work, and led us
to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be accurate for all three classifiers, in particular in regions with enough density
(i.e., from $\mathbf{x}=-3$ to $\mathbf{x}=2$) and where the capacity of the
classifier is focused. By contrast, in regions where the densities are low
(i.e., at the boundaries) the approximated ratio show to have high variance
since the underlying classifier cannot make very accurate predictions, because
of the few training samples it is trained on in these parts of the input space.
Accordingly, when leveraging the fact that densities are mixtures and decomposing their
ratio, the capacity of the underlying classifiers can be focused on easier
supervised learning tasks, resulting as expected in even more accurate approximations, as
shown in cyan for all three classifiers.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work well, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$
almost monotonic with the ratio. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
Similarly, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is also key to obtain accurate results. While standard
histograms with fixed bins have been used here for illustrative purposes, we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figures/fig3.pdf}
    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat g$.}
    \label{fig:3}
\end{figure}

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: g = 0$ versus the alternative $H_1: g > 0$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | g = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | g=\theta)} \; .
\end{equation}
Following the approach outlined in Section~\ref{sec:generalized-likelihood-ratio},
computing the generalized likelihood ratio statistic requires finding
the maximum likelihood estimator for the denominator, which can be carried out using
approximated likelihood ratios. Reusing the decomposed neural network-based
approximation and minimizing the mean negative log-ratio of some observed data
generated for $g=0.05$,
Figure~\ref{fig:3} shows the empirical distribution of $\hat g$ when repeating the generation and optimization process.
The figure clearly suggests that the proposed inference procedure yields in this case
an unbiased estimator ($\mathbb{E} [\hat g] = 0.04995$, as empirically estimated over $1000$ trials).
From there, the generalized likelihood ratio can be approximated as in the simple
hypothesis setting.

>>>>

\if1\blind
{
  \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
  \author{Kyle Cranmer$^1$, Juan Pavez$^2$ and Gilles Louppe$^1$\hspace{.2cm}\\
          $^1$New York University\\
          $^2$Federico Santa María University}
  \maketitle
} \fi

@@

As outlined above, any consistent probabilistic classification algorithm can be
used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
expected to smoothly vary around $\mathbf{x}$, we would however recommend
learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
around $\mathbf{x}$, such as neural networks. For small training sets, tree-based
methods are not expected to work so well for this use case, since they usually
model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
In such cases where $\hat s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased
approximation of the density ratio.
Provided enough training data, accurate results can however still be achieved,
given the universal approximator capacity of tree-based models.

@@

As a simple and illustrative example, let us first consider the approximation of
the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
Figure~\ref{fig:1a} and defined as
\begin{align}
p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \gamma) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \gamma) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
\end{align}
where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $\gamma$ is a parameter of $p(\mathbf{x}|\theta_0)$
set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
figure and will be used later for inference.

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:1b}, \ref{fig:1c} and \ref{fig:1d} show the approximated
ratio when using respectively a linear model, a 2-layer neural network and a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when both decomposing the mixtures and calibrating the underlying classifiers, as
described in Section~\ref{sec:mixture}.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
        \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
        \label{fig:1a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
        \caption{Linear model}
        \label{fig:1b}
    \end{subfigure}

@@

    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
        \caption{Neural network}
        \label{fig:1c}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
        \caption{Random forest}
        \label{fig:1d}
    \end{subfigure}
    \caption{Approximation of the log-likelihood ratio
             $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
             with calibrated classifiers.}
\end{figure}

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat
s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate estimates
of the ratio, even when using universal approximators such as a neural network
or a random forest. This observation is at the origin of this work, and led us
to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, in particular in regions with enough density
and where the capacity of the
classifier is focused. By contrast, in regions where the densities are low
(i.e., at the boundaries) the approximated ratio show to less acurate
since the underlying classifier cannot make precise predictions, because
of the few training samples it is trained on in these parts of the input space.
When leveraging the fact that densities are mixtures and decomposing their
ratio, the capacity of the underlying classifiers can be focused on easier
supervised learning tasks, resulting as expected in even more accurate approximations, as
shown in cyan.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis.

\begin{figure}
    \centering

    \begin{subfigure}[t]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2a.pdf}
        \caption{True MLEs vs. approximated MLEs.}
        \label{fig:2a}
    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[t]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figures/fig2b.pdf}
        \caption{$-2 \log \frac{p({\cal D}|\gamma=0.05)}{\sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)}$}
        \label{fig:2b}
    \end{subfigure}

    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$.}
    \label{fig:2}
\end{figure}

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
\end{equation}
Following the approach outlined in
Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
likelihood ratio statistic requires finding the maximum likelihood estimator
(MLE) for the denominator, which can be carried out using approximated
likelihood ratios. Reusing the decomposed neural network-based approximation and
minimizing the mean negative log-ratio of some observed data generated for
$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs
over resampled observed data,
compared to the distribution of the true MLEs. The figure clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating estimator bias.  Figure \ref{fig:2b} also confirms
that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
as should be the true  generalized likelihood ratio statistic.
